---
title: Surrogate Variable Analysis
author: "Siddharth Reed"
date: "`r Sys.Date()`"
lightbox: true
format:
  html:
    fig-width: 9
    fig-height: 5
    smooth-scroll: true
    toc: true
    number_sections: true
    code-fold: true
    theme:
        dark: darkly
        light: flatly
---
{{< include _imports.qmd >}}

Load SVs and sample data
```{r load_data}
all.sample.metadata <- 
    load_sample_metadata() %>%
    select(all_of(c('SampleID', RELEVANT_METADATA_COLUMNS)))
```

Load all SVs
```{r load_SV_data}
all.SVs.df <- 
    check_cached_results(
        results_file=here('results', 'all.SV.data.rds'),
        # force_redo=TRUE,
        results_fnc=
            function() {
                list_all_results_files_in_set(set.name='SVs') %>% 
                mutate(
                    SVs.df=
                        future_pmap(
                            .l=list(SVs.filepath),
                            .f=
                                function(SVs.filepath, sample.metadata) {
                                    SVs.filepath %>% 
                                    read_tsv(show_col_types=FALSE) %>% 
                                    left_join(
                                        sample.metadata,
                                        by=join_by(SampleID)
                                    )
                                },
                            sample.metadata=all.sample.metadata,
                            .progress=TRUE
                        )
                )
            }
    ) %>%
    mutate(
        SVs.estimated=
            pmap(
                .l=list(SVs.df),
                .f=~ .x %>% colnames() %>% str_detect('^SV') %>% sum()
            ) %>% 
            unlist(),
        z.score.counts=ifelse(z.score.counts, 'zscore -> SVs', 'unscaled -> SVs')
    ) %>% 
    select(-c(n.SVs, SVs.filepath))
```

# Correlation of SVs between sample sets

```{r compute_SV_SV_correlation}
# #| eval: false
compute_SV_correlations <- function(SVs.df.1, SVs.df.2, ...) {
    SVs.df.1 <- 
        SVs.df.1 %>% 
        as.data.frame() %>%
        column_to_rownames('SampleID') %>% 
        select(starts_with('SV'))
    SVs.df.2 <- 
        SVs.df.2 %>%
        as.data.frame() %>%
        column_to_rownames('SampleID') %>% 
        select(starts_with('SV'))
    common.samples <- 
        intersect(rownames(SVs.df.1), rownames(SVs.df.2))
    cor.mat <- 
        cor(
            SVs.df.1[common.samples, ],
            SVs.df.2[common.samples, ]
        ) %>%
        as.data.frame()
    rownames(cor.mat) <- colnames(SVs.df.1)
    colnames(cor.mat) <- colnames(SVs.df.2)
    cor.mat %>% 
    rownames_to_column('SVs.1') %>%
    as_tibble() %>%
    pivot_longer(
        -c(SVs.1),
        names_to='SVs.2',
        values_to='cor'
    )
}
compute_all_SV_correlations <- function(all.SVs.df){
    tribble(
        ~AD.definition.column, ~sample.strategy.1, ~sample.strategy.2, 
        'AD_CERAD_withDLB',    'All.Samples',      'All.Samples',
        'AD_CERAD_withDLB',    'All.Samples',      'AD.Samples',
        'AD_CERAD_withDLB',    'All.Samples',      'Control.Samples',
        'AD_CERAD_withDLB',    'AD.Samples',       'Control.Samples',
        # 'AD_CERAD_withDLB',    'No.Others',        'AD.Samples',
        # 'AD_CERAD_withDLB',    'No.Others',        'Control.Samples',
    ) %>%
    inner_join(
        all.SVs.df,
        relationship='many-to-many',
        by=join_by(AD.definition.column, sample.strategy.1 == sample.strategy)

    ) %>% 
    inner_join(
        all.SVs.df,
        suffix=c('.1', '.2'),
        relationship='many-to-many',
        by=
            join_by(
                CPM.cutoff, residual.model, 
                AD.definition.column,
                # SVs.est.strategy, SVs.estimated,
                z.score.counts, 
                sample.strategy.2 == sample.strategy
            )
    ) %>% 
    mutate(
        SV.corr=
            pmap(
                .l=.,
                .f=compute_SV_correlations,
            )
    ) %>% 
    unnest(SV.corr)
}
SV.correlations.df <- 
    check_cached_results(
        results_file=file.path(NOTEBOOK_CACHE_DIR, 'all.SV.data.tsv'),
        # force_redo=TRUE,
        results_fnc=compute_all_SV_correlations,
        all.SVs.df=all.SVs.df
    ) %>% 
    mutate(
        cor.label=round(cor, 2),
        comparison=
            case_when(
                sample.strategy.1 == sample.strategy.2 ~ sample.strategy.1,
                TRUE ~ glue('{sample.strategy.1} vs {sample.strategy.2}'),
            ),
        across(
            .cols=c(SVs.1, SVs.2),
            .fn=~ fct_reorder(.x, .x %>% str_remove('^SV') %>% as.integer())
        )
        # sv.int.1=SVs.1 %>% str_remove('^SV') %>% as.integer(),
        # SVs.1=fct_reorder(SVs.1, ),
        # sv.int.2=SVs.2 %>% str_remove('^SV') %>% as.integer(),
        # SVs.2=fct_reorder(SVs.2, sv.int.2)
    )
```

Here we plot the correlation of SVs estimated on different set of samples, specifically we compare SVs estimated from AD samples only or Controls only to All samples.
We only compare the SV correlations between common samples for each pair of different SV sets.

```{r plot_SV_SV_correlation}
#| echo: false
#| results: asis
#| fig-width: 15
#| fig-height: 15
# Now plot correlation of SVs across datasets
nesting.vars <- 
    c(
        'CPM.cutoff',
        'residual.model',
        'comparison',
        'z.score.counts'
    )
SV.correlations.df %>% 
    # # only compare SVs esstimated on the same sample sets
    # filter(comparison == sample.strategy.1) %>% 
    nest(plot.df=-all_of(nesting.vars)) %>% 
    mutate(
        test=pmap(.l=list(plot.df), .f=~ glue('{nrow(.x)}x{ncol(.x)}')),
        figures=
            pmap(
                .l=.,
                .f=plot_heatmap,
                facet.col='SVs.estimated.1',
                facet.row='SVs.estimated.2',
                x.var='SVs.1',
                y.var='SVs.2',
                fill.var='cor', 
                # label.var='cor.label',
                # label.size=2,
                # label.color='black',
                scales='free',
                space='free',
                # independent='all',
                fill.axis.label.accuracy=0.01,
                plot.elements=list(scale_fill_gradient2(low="red", mid="lightgrey", high="blue", midpoint=0)),
                axis.text.x=element_text(angle=45, vjust=1, hjust=1),
                axis.line=element_blank(),
                axis.title=element_blank(),
                .progress=TRUE
            )
    ) %>% 
        {.} -> tmp; ggsave(here('test.pdf'), tmp$figures[[1]])
    custom_render_tabset(
        tabset_vars=nesting.vars,
        output_var='figures'
        # output_var='test'
    )
    # render_tabset(
    #     # tabset_vars=nesting.vars,
    #     # heading_levels=c(1, rep(NA, length(nesting.vars))),
    #     # tabset_vars=c('title', nesting.vars),
    #     starting.hlevel=2,
    #     tabset_vars=nesting.vars,
    #     output_vars='figures'
    # )
```

# Correlation of SVs and Sample Metadata

```{r compute_SV_metadata_correlation}
# #| eval: false
make_correlation_data <- function(SVs.df, model.vars, ...) {
    # SVs.df=all.SVs.df$SVs.df[[1]]; model.vars=c('sex', 'RACE', 'Biobank', ALL_PHENOTYPE_COLUMNS)
    for (variable in model.vars) {
        is.constant <- SVs.df %>% pull(variable) %>% {all(duplicated(.)[-1L])}
        if (is.constant) {
            model.vars <- model.vars[model.vars != variable]
        }
    }
    all.vars <- 
        c(
            model.vars,
            grep('^SV', colnames(SVs.df), value=TRUE)
        )
    SVs.df %>% 
    canCorPairs(

        formula=sprintf('~ %s', paste(all.vars, collapse='+')) %>% formula(),

        data=.

    ) %>%
    as.data.frame() %>% 
    rownames_to_column('Var1') %>% 
    pivot_longer(
        -c(Var1),
        names_to='Var2',
        values_to='corr'
    )
}
compute_all_SV_metadata_correlations <- function(all.SVs.df, model.variables) {
    all.SVs.df %>% 
    mutate(
        corr.df=
            pmap(
                .l=.,
                .f=make_correlation_data,
                model.vars=model.variables
            )
    ) %>% 
    select(-SVs.df) %>% 
    unnest(corr.df)
}
factorize_for_plotting <- function(SV.correlations.df){
    all.vars <- 
        c(SV.correlations.df$Var1, SV.correlations.df$Var2) %>%
        unique() %>%
        {
            c(
                grep('^SV', ., value=TRUE, invert=TRUE) %>% sort(),
                grep('^SV', ., value=TRUE) %>% {.[as.integer(str_remove(., 'SV'))]}
            )
        }
    SV.correlations.df %>% 
    mutate(across(c(Var1, Var2), ~ factor(.x, levels=all.vars))) %>% 
    mutate(
        Var1.ordered=
            case_when(
                as.integer(Var1) <= as.integer(Var2) ~ Var1,
                as.integer(Var1) >  as.integer(Var2) ~ Var2
            ),
        Var2.ordered=
            case_when(
                as.integer(Var1) <= as.integer(Var2) ~ Var2,
                as.integer(Var1) >  as.integer(Var2) ~ Var1
            ),
        across(
            .cols=c(Var1, Var2),
            .names='{.col}.label',
            .f=
                ~ case_when(
                    grepl('^SV', .x) ~ 'SV',
                    .x %in% all.vars ~ 'phenotype'
                ) %>% 
                factor()
        ),
        var.comparison=
            case_when(
                as.integer(Var1.label)  > as.integer(Var2.label) ~ glue('{Var1.label} vs {Var2.label}'),
                as.integer(Var1.label)  < as.integer(Var2.label) ~ glue('{Var2.label} vs {Var1.label}'),
                as.integer(Var1.label) == as.integer(Var2.label) ~ glue('within {Var1.label}'),
                .default='???'
            )
    ) %>%
    select(-c(Var1, Var2)) %>%
    dplyr::rename_with(~ str_remove(.x, '.ordered')) %>%
    select(-c(ends_with('.label')))
}
SV.metadata.correlations.df <- 
    check_cached_results(
        results_file=file.path(NOTEBOOK_CACHE_DIR, 'all.SV.metadata.correlations.tsv'),
        # force_redo=TRUE,
        results_fnc=compute_all_SV_metadata_correlations,
        all.SVs.df=all.SVs.df
    ) %>% 
    filter(SVs.estimated == NUM_SVS_TO_GENERATE) %>% 
    factorize_for_plotting()
```

Here we plot the correlation of estimated SVs with various sample characteristics e.g. AD phenotypes, Age and model co-variates. We previously showed that 

```{r plot_SV_metadata_correlation}
# #| eval: false
#| echo: false
#| results: asis
#| output: asis 
#| fig-width: 15
#| fig-height: 15

nesting.vars <- 
    c(
        'CPM.cutoff',
        'residual.model',
        'AD.definition.column',
        'sample.strategy',
        'z.score.counts'
    )
SV.metadata.correlations.df %>% 
    mutate(corr.label=round(corr, 2)) %>% 
    filter(SVs.estimated == NUM_SVS_TO_GENERATE) %>% 
    # nest(plot.df=-all_of(c(nesting.vars, 'residual.model', 'sample.strategy'))) %>% 
    nest(plot.df=-all_of(c(nesting.vars))) %>% 
    mutate(
        figures=
            pmap(
                .l=.,
                .f=plot_heatmap,
                # facet.group='SVs.estimated', facet.ncol=7,
                # facet.col='sample.strategy', facet.row='residual.model'
                facet.col='var.comparison',
                y.var='Var2',
                x.var='Var1',
                fill.var='corr', 
                label.var='corr.label',
                label.size=2,
                label.color='white',
                scales='free',
                # space='free',
                independent='all',
                axis.text.x=element_text(angle=45, vjust=1, hjust=1),
                axis.line=element_blank(),
                axis.title=element_blank(),
                .progress=TRUE
            )
    ) %>% 
    add_column(title='Correlation of SVs and Sample Metadata') %>% 
    render_tabset(
        # heading_levels=c(1, rep(NA, length(nesting.vars))),
        # tabset_vars=c('title', nesting.vars),
        starting.hlevel=2,
        tabset_vars=nesting.vars,
        output_vars='figures'
    )
```

# variancePartition Analysis

## How much does iteratively regressing how SVs affect total peak vairance? {#sec-varianceElbow}

First we want to show the difference in total variance of the input matrix $R$ before and after iteratively regressing out up to $N$ SVs. 
For each SV regressed matrix $$R_i = residuals(R, SV_{1 \dots i}) \quad | \quad \forall i \in \{1\dots N\}$$ we calculate the variance of each peak across all samples, then we compute summary statiscs across all peak variances. 
Here I am plotting the mean and total peak variance across all SV-residualzied matrices $R_{1 \dots N} = \{R_1, \dots, R_N\}$, as we can see there is an elbow like decrease in peak variance as we regress out more SVs.

```{r load_peak_variances}
#| echo: false
# #| eval: false

summarize_peak_variances <- function(
    peak.stats.filepath,
    ...){
    peak.stats.filepath %>%
    read_tsv(show_col_types=FALSE) %>%
    group_by(across(-c(PeakID, peak.variances))) %>% 
    summarize(
        .groups='drop',
        across(
            .cols=c(peak.variances),
            .names='stat.{.fn}',
            .fns=STAT_SUMMARY_FNCS_LIST
        )
    ) %>%
    pivot_longer(
        starts_with('stat.'),
        names_to='stat',
        names_prefix='stat.',
        values_to='value'
    )
}
all.peak.variance.results.df <- 
    check_cached_results(
        results_file=file.path(NOTEBOOK_CACHE_DIR, 'peak.variance.data.tsv'),
        force_redo=FORCE_REDO,
        # force_redo=TRUE,
        results_fnc=
            function() {
                list_all_results_files_in_set(set.name='peak.variance.stats') %>%
                mutate(
                    elbow=
                        future_pmap(
                            .l=.,
                            .f=summarize_peak_variances,
                            .progress=TRUE
                        )
                ) %>% 
                unnest(elbow)
            }
    ) %>% 
    filter(
        z.score.counts,
        n.SVs == NUM_SVS_TO_GENERATE
    ) %>% 
    dplyr::rename('SVs.Regressed.Out'=SVs.included) %>% 
    select(
        CPM.cutoff, residual.model, AD.definition.column, sample.strategy,
        z.score.counts, n.SVs, SVs.Regressed.Out,
        stat, value
    )
```

```{r plot_peak_variances_elbow, results='asis'}
#| echo: false
#| fig-height: 4
#| fig-width: 10
# #| eval: false

nesting.vars <- 
    c(
        'CPM.cutoff',
        'residual.model'
        # 'sample.strategy'
    )
all.peak.variance.results.df %>% 
    filter(stat %in% c('mean', 'total')) %>% 
    mutate(stat=glue('{stat} variance across peaks')) %>% 
    order_variable_for_plotting() %>% 
    nest(plot.df=-all_of(nesting.vars)) %>% 
    mutate(
        figures=
            pmap(
                .l=.,
                # .f=plot_ribbonplot, ymin.var='ymin', ymax.var='ymax',
                .f=plot_lineplot, 
                x.var='SVs.Regressed.Out',
                y.var='value',
                size=1, linewidth=0.75,
                group.var='sample.strategy', color.var='sample.strategy', facet.col='stat',
                scales='free_y', space='fixed', independent='y',
                axis.title.y=element_blank(),
                # plot.elements=list(geom_hline(yintercept=0.3, linewidth=0.15, linetype='dashed')),
                .progress=TRUE
            )
    ) %>% 
    custom_render_tabset(
        starting.hlevel=3,
        tabset_vars=nesting.vars,
        output_var='figures'
    )
```

## How much peak variance do SVs capture?
# CRD Elbow Plot of LEFs vs SVs

First load all the data where we calcualted all decorate CRD LEFs from separate matrices with different sets of SVs regressed out i.e. {1}, {1,2}, {1,2,3,...N} etc. 
Most matrices only had 1-4 SVs estimated.
```{r load_elbow_data}
#| eval: false
#| cache: true
all.elbow.data.df <- 
    ELBOW_RESULTS_DIR %>% 
    parse_results_filelist(
        suffix='elbow.plot.data.tsv',
        filename.column.name='filename'
    ) %>%
    dplyr::rename('elbow.filepath'=filepath) %>% 
    select(-c(filename)) %>%
    {read_tsv(.$elbow.filepath)} %>% count(SVs.included, meanClusterSize)
```
Here we build and elbow plot with showing the LEF distribution across peaks for each residual matrix with the specified SVs regressed out. This shows us 
```{r plot_LEF_elbow}
#| output: asis
#| echo: false
#| eval: false

```

