---
title: Surrogate Variable Analysis
author: "Siddharth Reed"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
lightbox: true
format:
  html:
    fig-width: 9
    fig-height: 5
    code-fold: true
    theme:
        dark: darkly
        light: flatly
---
{{< include _imports.qmd >}}

Load SVs and sample data
```{r load_data}
source(here('scripts', 'utils.decorate.R'))
all.sample.metadata <- 
    load_sample_metadata()
all.SVs.df <- 
    list_all_SVs() %>%
    mutate(
        SVs.df=
            pmap(
                .l=list(SVs.filepath),
                .f=
                    function(SVs.filepath, sample.metadata) {
                        SVs.filepath %>% 
                        read_tsv(show_col_types=FALSE) %>% 
                        left_join(
                            sample.metadata,
                            by=join_by(SampleID)
                        )
                    },
                sample.metadata=all.sample.metadata
            )
    ) %>% 
    select(-c(SVs.filepath))
```
Functions 
```{r plot_functions}
make_correlation_data <- function(SVs.df, model.vars, ...) {
    # SVs.df=all.SVs.df$SVs.df[[1]]; model.vars=c('sex', 'RACE', 'Biobank', ALL_PHENOTYPE_COLUMNS)
    for (variable in model.vars) {
        is.constant <- SVs.df %>% pull(variable) %>% {all(duplicated(.)[-1L])}
        if (is.constant) {
            model.vars <- model.vars[model.vars != variable]
        }
    }
    all.vars <- 
        c(
            model.vars,
            grep('^SV', colnames(SVs.df), value=TRUE)
        )
    SVs.df %>% 
    canCorPairs(
        formula=sprintf('~ %s', paste(all.vars, collapse='+')) %>% formula(),
        data=.
    ) %>%
    as.data.frame() %>% 
    rownames_to_column('Var1') %>% 
    pivot_longer(
        -c(Var1),
        names_to='Var2',
        values_to='corr'
    ) %>% 
    mutate(across(c(Var1, Var2), ~ factor(.x, levels=all.vars)))
}
```

# Correlation of SVs between sample sets

Here we plot the correlation of SVs estimated on different set of samples, specifically we compare SVs estimated from AD samples only or Controls only to All samples.
We only compare the SV correlations between common samples for each pair of different SV sets.
```{r plot_SV_SV_correlation}
#| output: asis 
#| echo: false
# #| fig-width: 11
# #| fig-height: 6
SV.correlations.df <- 
    tribble(
        ~sample.set.1, ~sample.set.2, 
        'All.Samples', 'AD.Samples',
        'All.Samples', 'Control.Samples',
        'No.Others',   'AD.Samples',
        'No.Others',   'Control.Samples',
    ) %>%
    left_join(
        all.SVs.df,
        by=join_by(sample.set.1 == sample.strategy)

    ) %>% 
    left_join(
        all.SVs.df,
        suffix=c('.1', '.2'),
        by=
            join_by(
                CPM.cutoff, residual.model, AD.definition.column,
                sample.set.2 == sample.strategy
            )
    ) %>% 
    mutate(
        SV.corr=
            pmap(
                .l=.,
                .f=
                    function(SVs.df.1, SVs.df.2, ...) {
                        SVs.df.1 <- 
                            SVs.df.1 %>% 
                            as.data.frame() %>%
                            column_to_rownames('SampleID') %>% 
                            select(starts_with('SV'))
                        SVs.df.2 <- 
                            SVs.df.2 %>%
                            as.data.frame() %>%
                            column_to_rownames('SampleID') %>% 
                            select(starts_with('SV'))
                        common.samples <- 
                            intersect(rownames(SVs.df.1), rownames(SVs.df.2))
                        cor.mat <- 
                            cor(
                                SVs.df.1[common.samples, ],
                                SVs.df.2[common.samples, ]
                            ) %>%
                            as.data.frame()
                        rownames(cor.mat) <- colnames(SVs.df.1)
                        colnames(cor.mat) <- colnames(SVs.df.2)
                        cor.mat %>% 
                        rownames_to_column('SVs.1') %>%
                        as_tibble() %>%
                        pivot_longer(
                            -c(SVs.1),
                            names_to='SVs.2',
                            values_to='cor'
                        )
                    }
            )
    ) %>% 
    select(-c(SVs.df.1, SVs.df.2)) %>%
    unnest(SV.corr) %>% 
    mutate(
        comparison=glue('{sample.set.1} vs {sample.set.2}'),
        cor.label=round(cor, 2),
        sv.int.1=SVs.1 %>% str_remove('^SV') %>% as.integer(),
        SVs.1=fct_reorder(SVs.1, sv.int.1),
        sv.int.2=SVs.2 %>% str_remove('^SV') %>% as.integer(),
        SVs.2=fct_reorder(SVs.2, sv.int.2)
    )
# Now plot correlation of SVs across datasets
SV.correlations.df %>% 
    make_nested_plot_tabs(
        max.header.lvl=2,
        # group.cols=c('CPM.cutoff', 'residual.model', 'comparison'),
        # group.cols=c('CPM.cutoff', 'residual.model'), facet.row='sample.set.2', facet.col='sample.set.1',
        # group.cols=c('CPM.cutoff', 'sample.set.1', 'residual.model'), facet.col='sample.set.2',
        group.cols=c('CPM.cutoff', 'sample.set.1'), facet.col='residual.model', facet.row='sample.set.2',
        plot.fnc=plot_heatmap,
        x.var='SVs.1',
        y.var='SVs.2',
        fill.var='cor', 
        label.var='cor.label',
        label.size=2,
        label.color='black',
        scales='free',
        space='free',
        independent=TRUE,
        fill.axis.label.accuracy=0.01,
        plot.elements=list(scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0)),
        axis.text.x=element_text(angle=45, vjust=1, hjust=1),
        axis.line=element_blank(),
        axis.title=element_blank()
    )
```

# Correlation of SVs and Sample Metadata

Here we plot the correlation of estimated SVs with various sample characteristics e.g. AD phenotypes, Age and model co-variates.
```{r plotSV_metadata_correlation}
#| output: asis 
#| echo: false
all.SVs.df %>% 
    mutate(
        corr.df=
            pmap(
                .l=.,
                .f=make_correlation_data,
                model.vars=
                    c(
                        'RACE',
                        'sex',
                        'Biobank',
                        ALL_PHENOTYPE_COLUMNS
                    )
            )
    ) %>% 
    select(-SVs.df) %>% 
        # {.} -> tmp; tmp; tmp$corr.df[[1]]
    unnest(corr.df) %>% 
    mutate(corr.label=round(corr, 2)) %>% 
    make_nested_plot_tabs(
        max.header.lvl=2,
        # group.cols=c('CPM.cutoff', 'residual.model'), facet.col='sample.strategy',
        # group.cols=c('residual.model', 'sample.strategy'),
        group.cols=c('CPM.cutoff', 'sample.strategy'), facet.col='residual.model',
        plot.fnc=plot_heatmap,
        y.var='Var1',
        x.var='Var2',
        fill.var='corr', 
        label.var='corr.label',
        label.size=2,
        label.color='white',
        # scales='free',
        # space='free',
        # independent=TRUE,
        # scales='free',
        axis.text.x=element_text(angle=45, vjust=1, hjust=1),
        axis.line=element_blank(),
        axis.title=element_blank()
    )
```

# variancePartition Analysis

What is the variance explained by the model variables pre-post residualizing out SVs?
How much varaince to the SVs account for in terms of chromatin accessability?

# CRD Elbow Plot of LEFs vs SVs

First load all the data where we calcualted all decorate CRD LEFs from separate matrices with different sets of SVs regressed out i.e. {1}, {1,2}, {1,2,3,...N} etc. 
Most matrices only had 1-4 SVs estimated.
```{r load_elbow_data}

    SVA_RESULTS_DIR %>% 
    parse_results_filelist(
        suffix='elbow.plot.data.tsv',
        filename.column.name='filename'
    ) %>%
    dplyr::rename('elbow.filepath'=filepath) %>% 
    select(-c(filename)) %>%
    {read_tsv(.$elbow.filepath)} %>% count(SVs.included, meanClusterSize)
```
Here we build and elbow plot with showing the LEF distribution across peaks for each residual matrix with the specified SVs regressed out. This shows us 
```{r plot_LEF_elbow}
#| output: asis

```

