title: CRD Explorations 
author: "Siddharth Reed"
date: "`r Sys.Date()`"
lightbox: true
format:
  html:
    fig-width: 9
    fig-height: 5
    smooth-scroll: true
    toc: true
    number_sections: true
    code-fold: true
    theme:
        dark: darkly
        light: flatly
---

{{< include _imports.qmd >}}

Load sample metadata
```{r load_metadata}
plan(multisession, workers=TOTAL_CORES)
NOTEBOOK_CACHE_DIR <- 
    here('results', 'notebook.cache', '02_CRD.Investigation')
stat.summary.fncs.list <- 
    c(
        'min'=min, 
        'q05'=~ quantile(.x, probs=0.05),
        'q25'=~ quantile(.x, probs=0.25),
        'mean'=mean, 'med'=median, 
        'q75'=~ quantile(.x, probs=0.75),
        'q95'=~ quantile(.x, probs=0.95),
        'max'=max,
        'var'=var
    )
all.sample.metadata <- 
    load_sample_metadata() %>%
    select(all_of(c('SampleID', RELEVANT_METADATA_COLUMNS)))
```
Load all SVs
```{r load_CRDs}
all.CRDs.df <- 
    check_cached_results(
        results_file=here('results', 'all.SV.data.rds'),
        # force_redo=TRUE,
        results_fnc=
            function() {
                list_all_results_files_in_set(set.name='') %>% 
                mutate(
                    SVs.df=
                        future_pmap(
                            .l=list(SVs.filepath),
                            .f=read_tsv, show_col_types=FALSE,
                            .progress=TRUE
                        )
                )
            }
    )
```

# Compare CRD features across subsets

# Compare effects of filtering/collapsing CRDs 

# Compare differential OCRs across CRDs across conditions

# Define AD/Control-specific CRDs and stratify plots by them
